ICT and Future Productivity:\\Evidence and Theory of a GPT\thanks{Correspondence: Department of Economics, Boston College, 140 Commonwealth Avenue, Chestnut Hill, MA 02467. Email: brianti@bc.edu (Marco Brianti) and gati@bc.edu (Laura G\'ati).}

### 2018-07-29 Latex/Draft_Summer2018/Summer2018.tex
read the last part of the paper
will read the whole again with fresh eyes tomorrow

We now turn to describing the parameter values and testing the empirical performance of the model. Following \cite{christiano2005nominal}, we partition the model parameters into two groups. We calibrate parameters of the first group using steady-state relations or results from previous works. We estimate the second group by minimizing the distance between the model-implied impulse responses and the empirical results presented in Section \ref{section:mainSpec}. In particular, the second group focuses on parameters specifically related to the effect of ICT investment on future TFP: (i) the standard deviation $\sigma_{\iota}$ of ICT shock $\zeta_t$, (ii) persistence $\rho_{\iota}$ of ICT shock $\zeta_t$ and (iii) the size of the spillover effect of ICT capital on productivity $\gamma$.

@@

We calibrate the model at a quarterly frequency. We set the discount factor $\beta$ to $0.97^{\frac{1}{4}}$ to match the average yield of $3$\% of the 3-month Treasury bill using data from 1989:Q1 to 2017:Q1. We set the ICT share $b$ equal to $0.0310$ following \cite{oulton2012long} and hard capital share $a$ equal to $0.3$. Moreover, adapting \cite{oulton2012long} accordingly to our additional assumptions, we derive steady state growth rates of final-sector technology $\Gamma^c$ and ICT-sector technology $\Gamma^i$ from solving the system,
\begin{equation}\label{equation:system_param}
\begin{cases}
 \frac{p_t}{p_{t-1}}  = \frac{\Gamma^c}{\Gamma^i} \\
 \frac{c_t}{c_{t-1}}  = (\Gamma^c)^{\frac{1-b-\gamma}{1-a-b-\gamma}}  (\Gamma^i)^{\frac{b+\gamma}{1-a-b-\gamma}}
\end{cases}
\end{equation}
where $p_t/p_{t-1}$ and $c_t/c_{t-1}$ are derived from the simple mean of their empirical related values using quarterly data from 1989:Q1 to 2017:Q1. This implies $\Gamma^c = 1.0034$ and $\Gamma^i = 1.0160$.\footnote{Notice that the value of $\gamma$ is unknown at this step since it is going to be evaluated using impulse-response matching. This implies that the solution for $\Gamma^c$, $\Gamma^i$, and $\gamma$ should be a fixed point where given $\gamma^*$ we obtain $(\Gamma^i)^*$ and $(\Gamma^c)^*$ solving \ref{equation:system_param}. Given $(\Gamma^i)^*$ and $(\Gamma^c)^*$ we obtain $\gamma^*$ by minimizing the impulse-response matching. At this step we are simply solving the system for $\gamma$ equal to zero but a more refine solution will be provided in future.} Finally, to avoid an over-identification problem we set the growth rate of sector-neutral technology $\Gamma^{\eta}$ equal to one.

@@

In this section, we estimate model’s key parameters governing the effect of ICT investment on future TFP by imposing theoretical impulse responses of an ICT shock to be as close as possible to the empirical ones. To enhance movement in ICT investment, we use a shock to the ICT-specific technological component $\zeta_t$. Using empirical impulse responses of TFP, ICT investment, consumption, and relative prices, we estimate four parameters: the variance of an ICT technology shock, $\sigma_{\iota}^2$, the persistence of the same shock, $\rho_{\iota}$, and the size of the spillover effect of ICT capital on TFP, $\gamma$. We estimate $\Omega = (\gamma, \ \sigma_{\iota}^2, \ \rho_{\iota})$ by minimizing,
\begin{eqnarray}\label{equation:min_prob_IRmatching}
\min_{\Omega} \big[  \hat{\Psi} - \Psi(\Omega)  \big]' \Lambda \big[  \hat{\Psi} - \Psi(\Omega)  \big]
\end{eqnarray}
where $\Psi(\Omega)$ denotes the mapping from $\Omega$ to the theoretical impulse responses, and $\hat{Psi}$ denotes the empirical impulse responses of an ICT shock to TFP, ICTI, consumption, and relative prices shown in Appendix \ref{section:mainSetResults}. Moreover, $\Lambda$ is a diagonal matrix which embodies variances of the empirical impulse responses. These elements serve as weights to impose relatively more precision on some estimates on the optimization problem. Finally, accordingly to what presented above, we use a horizon of $40$ periods for each response which sum up to a total of $160$ elements for three theoretical parameters only.

Table \ref{table:estimated_parameters} shows results of Problem \ref{equation:min_prob_IRmatching}. The estimate of the spillover effect of ICT capital on TFP, $\gamma$, is $0.58$ indicating that empirical responses favor a causality relation from an ICT capital deepening to total factor productivity. Moreover, estimates of variance and persistence of an ICT-specific technological shock $\sigma^2_{\iota}$ and $\rho_{\iota}$ are $0.01^2$ and $0.9$, respectively. Results imply both a remarkable size and persistence of an ICT sector-specific technological shock $\zeta_t$ in the economy. 
% does it make sense? I dont know what to say for sigmaiota and rhoiota...

Using the estimated parameter values in Table \ref{table:estimated_parameters}, Figure \ref{fig:IR_matching} displays empirical impulse responses along with theoretical ones. Our simple model is able to mimic empirical movements in TFP, ICT investment, consumption, and relative prices. In both the model and the data, a sector-specific technological shock $\zeta_t$ has a delayed positive effect on TFP which tend to last over the medium and long run. Moreover, in both cases, ICT investment has a large initial peak and then tends to remain persistently positive over the $40$-quarter horizon. In the model and the data, consumption responds positively on impact, reaches its peak after $4$ years and persistently remains positive. Finally, model-generated and empirical impulse responses of relative prices have a large and negative impact response which dies out in about $6$ years.

@@

Although both the model and impulse-response matching exercise should be considered as preliminary, we strongly believe they are already able to rationalize empirical results obtained in Section \ref{section:empirics} and convey the main message of this paper. Formally, the question we tried to answer is why a contemporaneous technological shock specific to the ICT sector is able to trigger increases in future total factor productivity. In other words, why an ICT-investment shock is able to increase productivity of ICT-good users with some lags.

We interpret ICT capital as being the general-purpose technology of the U.S. economy over the last $3$ decades. Specifically, when an ICT technology shock hits the economy, both sectors invest more on ICT capital goods since they are now cheaper to purchase and easier to produce. Now, in order to link current higher investment on ICT durable goods to future TFP, we assume that both production functions are augmented by a spillover effect related to the overall stock of ICT capital which is not internalized by the agents.\footnote{This externality is governed by the parameter $\gamma$ which has been estimated via impulse-response matching to be robustly larger than zero. Using different empirical results, changing the initial calibration, or the methodology do not qualitative change the estimated value of $\gamma$ close to $0.6$.} These assumptions are mechanically suitable for two main reasons. First of all, this spillover enters into the estimated Solow residual since it takes the form of an externality which implies that industries are not paying for this additional benefit. Moreover, since this spillover arises from a state variable, it leads to a dynamics where the accumulation of ICT durable goods is a slow process and the benefits of an ICT technology shock show up in the productivity of ICT-capital users with some lags and this effect tends to be persistent.

Although this assumption is mathematically suitable to track empirical impulse responses, we consider important to highlight that this positive externality is sustained by an intuitive and reliable economic intuition which distinguishes the general-purpose nature of ICT capital. Since the purpose of ICT goods is to enhance communication, the quality and speed of information sharing is also determined by the overall diffusion of these goods. As a simple example, assume that in 1995 a scholar decides to purchase a computer with an Internet subscription in order to speed up the acquisition of information. She is very satisfied of this investment and does not regret the paid amount. However, at the same time, a former colleague who is now working in a different university decides to buy a computer with an Internet subscription as well. Once she learns of this development, she also decides to exploit her computer to exchange research material with her former colleague. They eventually decide to start a new project together since being in two different universities is not anymore a constraint for the quality of their joint research. Both scholars did not pay for this additional benefit since they did not expect to have simultaneously a computer with an Internet subscription. However, what is actually happening is that their research production has been boosted by the positive externality that also the other colleague is endowed with the same technology.

As a result, we expect that the nature of ICT capital should display two main features. First of all, the effectiveness of an ICT good must be related to its own diffusion as pointed out by the spillover effect. Moreover, the cost of ICT items should not initially reflect an unexpected increase in their diffusion as pointed out by the assumption that the first feature is not internalized.

@@

In the first part of the paper, we investigate the effect of technological shocks specific to the ICT sector on future TFP. Using Structural VAR techniques on quarterly aggregate U.S. data from 1989, we impose two identifying restrictions to correctly identify ICT shocks: (i) they must be orthogonal to current productivity and (ii) maximize the impact response of ICT investment. In response to an ICT shock, ICT investment increases on impact and stays significant for more than 10 years; relative prices persistently decline as an additional support to the reliability of our identification strategy; and TFP rises after three quarters and remains significant for more than 10-years. In addition, variance decomposition analysis suggests that the quantitative impact of ICT shocks on TFP is economically remarkable since they explain about one third of the overall TFP fluctuations over a 10-year horizon.

In order to check whether our identification assumptions may confound our shock with a news shock which contemporaneously boosts ICT investment, we provide a series of different identification strategies which deliver analogous results. Our main robustness check is a sequential identification of news shocks first and then ICT shocks which encouragingly does not affect estimates. We view this as confirmation of the causality relation which goes from current ICT investment to future TFP.

In the second part of the paper, we rationalize our empirical results by analyze a two-sector DSGE model \`a la \cite{greenwood1997long} augmented by two main assumptions: (i) the sector with the faster technological improvement produces specifically ICT capital, and (ii) both production functions are augmented with a spillover related to the overall level of ICT capital which is not internalized by the agents. In particular, this last assumption points out our interpretation of the general-purpose nature of ICT capital: the effectiveness of ICT capital is enhanced by its diffusion and its cost does not reflect this indirect benefit. We use both empirical and theoretical results to estimate main parameters of the model via impulse-response matching to an ICT technology shock. Results point out that the parameter which governs the size of the spillover effect is positive and large, suggesting that our interpretation of the general-purpose nature of ICT capital is confirmed by the data.

>>>>

We now turn to describing the parameter values and testing the empirical performance of the model. Following \cite{christiano2005nominal}, we partition the model parameters into two groups. We calibrate parameters of the first group using steady-state relations or results from previous works. We estimate the second group by minimizing the distance between the model-implied impulse responses and the empirical results presented in Section \ref{section:mainSpec}. In particular, the second group focuses on parameters specifically related to the effect of ICT investment on future TFP: (i) the standard deviation $\sigma_{\iota}$ of the ICT shock $\zeta_t$, (ii) the persistence $\rho_{\iota}$ of the ICT shock $\zeta_t$ and (iii) the size of the spillover effect of ICT capital on productivity $\gamma$.

@@

We calibrate the model at a quarterly frequency. We set the discount factor $\beta$ to $0.97^{\frac{1}{4}}$ to match the average yield of $3$\% of the 3-month Treasury bill using data from 1989:Q1 to 2017:Q1. We set the ICT share $b$ equal to $0.0310$ following \cite{oulton2012long} and hard capital share $a$ equal to $0.3$. Moreover, adapting \cite{oulton2012long} accordingly to our additional assumptions, we derive steady state growth rates of final-sector technology $\Gamma^c$ and ICT-sector technology $\Gamma^i$ from solving the system,
\begin{equation}\label{equation:system_param}
\begin{cases}
 \frac{p_t}{p_{t-1}}  = \frac{\Gamma^c}{\Gamma^i} \\
 \frac{c_t}{c_{t-1}}  = (\Gamma^c)^{\frac{1-b-\gamma}{1-a-b-\gamma}}  (\Gamma^i)^{\frac{b+\gamma}{1-a-b-\gamma}}
\end{cases}
\end{equation}
where $p_t/p_{t-1}$ and $c_t/c_{t-1}$ are derived from the simple mean of their empirical related values using quarterly data from 1989:Q1 to 2017:Q1. This implies $\Gamma^c = 1.0034$ and $\Gamma^i = 1.0160$.\footnote{Notice that the value of $\gamma$ is unknown at this step since it is going to be evaluated using impulse-response matching. This implies that the solution for $\Gamma^c$, $\Gamma^i$, and $\gamma$ should be a fixed point where given $\gamma^*$ we obtain $(\Gamma^i)^*$ and $(\Gamma^c)^*$ solving \ref{equation:system_param}. Given $(\Gamma^i)^*$ and $(\Gamma^c)^*$ we obtain $\gamma^*$ by minimizing the impulse-response matching. At this step, however, we are simply solving the system assuming $\gamma$ equal to zero, but a more refined solution will be provided in future.} Finally, to avoid an over-identification problem we set the growth rate of sector-neutral technology $\Gamma^{\eta}$ equal to one.

@@

In this section, we estimate the model's key parameters governing the effect of ICT investment on future TFP by imposing theoretical impulse responses of an ICT shock to be as close as possible to the empirical ones. To enhance movement in ICT investment, we use a shock to the ICT-specific technological component $\zeta_t$. Using empirical impulse responses of TFP, ICT investment, consumption, and relative prices, we estimate three parameters: the variance of an ICT technology shock, $\sigma_{\iota}^2$, the persistence of the same shock, $\rho_{\iota}$, and the size of the spillover effect of ICT capital on TFP, $\gamma$. We estimate $\Omega = (\gamma, \ \sigma_{\iota}^2, \ \rho_{\iota})$ by minimizing,
\begin{eqnarray}\label{equation:min_prob_IRmatching}
\min_{\Omega} \big[  \hat{\Psi} - \Psi(\Omega)  \big]' \Lambda \big[  \hat{\Psi} - \Psi(\Omega)  \big]
\end{eqnarray}
where $\Psi(\Omega)$ denotes the mapping from $\Omega$ to the theoretical impulse responses, and $\hat{\Psi}$ denotes the empirical impulse responses of an ICT shock to TFP, ICTI, consumption, and relative prices shown in Appendix \ref{section:mainSetResults}. $\Lambda$ is a diagonal weighting matrix, which is the inverse of the variance matrix of the empirical impulse responses. These elements serve as weights to impose more precision on some estimates of the optimization problem. Finally, consistent with our choice of horizons presented above, we use a horizon of $40$ periods for each response. % which sum up to a total of $160$ elements for three theoretical parameters only.

Table \ref{table:estimated_parameters} shows results of Problem \ref{equation:min_prob_IRmatching}. The estimate of the spillover effect of ICT capital on TFP, $\gamma$, is $0.58$, while estimates of variance and persistence of an ICT-specific technological shock $\sigma^2_{\iota}$ and $\rho_{\iota}$ are $0.01$ and $0.9$ respectively.\footnote{Estimated parameter values are strongly robust to various alternative specifications. Using different empirical results, changing the initial calibration, the number of parameters matched or the methodology do not qualitative change the estimated value of $\gamma$ close to $0.6$.} These results convey a number of interesting messages regarding the fit of our two-sector model. Most importantly, the key question of our model was whether the presence of ICT shocks alone can account for the empirical responses we observe in data, or whether ICT diffusion plays a crucial role in this dynamics. This amounts to asking whether $\gamma = 0$ is supported by the data. Clearly, our estimate of $0.58$ rejects this hypothesis strongly. Thus, the data favor an interpretation of our model in which there are considerable spillovers to ICT capital. Accordingly, ICT shocks do not need to be large in magnitude to have sizable effects on TFP, GDP and consumption. This is indicated by the relatively small number we obtain for $\sigma^2_{\iota}$ of $0.01$. Confronted with data, the model is clearly telling a story in which there are small, but persistent ($\rho_{\iota} = 0.9$) ICT shocks that lead considerable medium-run TFP growth because of large spillovers from ICT capital accumulation.
% does it make sense? I dont know what to say for sigmaiota and rhoiota...

Using the estimated parameter values in Table \ref{table:estimated_parameters}, Figure \ref{fig:IR_matching} displays empirical impulse responses along with theoretical ones. Our simple model is able to mimic empirical movements in TFP, ICT investment, consumption, and relative prices. In both the model and the data, a sector-specific technological shock $\zeta_t$ has a delayed positive effect on TFP which tends to last over the medium and long run. Moreover, in both cases, ICT investment has a large initial peak and then tends to remain persistently positive over the $40$-quarter horizon. In the model and the data, consumption responds positively on impact, reaches its peak after $4$ years and persistently remains positive. Finally, model-generated and empirical impulse responses of relative prices have a large and negative impact response which dies out in about $6$ years.

@@

Although both the model and impulse-response matching exercise should be considered as preliminary, we believe that they are already able to rationalize the empirical results obtained in Section \ref{section:empirics} and thus convey the main message of this paper. Formally, the question we tried to answer is why a contemporaneous technological shock specific to the ICT sector is able to trigger increases in future total factor productivity even when it is too small to affect TFP today. In other words, why is an ICT-investment shock able to increase the productivity of ICT-good users with lags?

We interpret ICT capital as being the general-purpose technology of the U.S. economy over the last three decades. Specifically, when an ICT technology shock hits the economy, both sectors invest more on ICT capital goods since they are now cheaper to purchase and easier to produce. Now, in order to link current higher investment in ICT goods to future TFP, we assume that the general-purpose nature of ICT goods manifests itself in the form of a spillover on both production functions which is not internalized by the agents.\footnote{This externality is governed by the parameter $\gamma$ which has been estimated via impulse-response matching to be robustly larger than zero.} Since the spillover comes from the overall stock of ICT capital, we interpret it as the diffusion of ICT capital.

This assumption is legitimate from a mathematical standpoint for two main reasons. First of all, the spillover enters the estimated Solow residual since, being an externality, it is a non-renumerated input. Additionally, since the spillover arises from a state variable, it leads to a dynamics where the benefits of an ICT technology shock show up in the productivity of ICT-capital users with some lags because the accumulation of ICT goods is a slow process. Thus, the spillover effect only comes into play only gradually once enough ICT capital is being accumulated. The slow-moving nature of the state variable capital also means that this effect tends to be persistent, which also helps us match the empirical responses we see in the data.\footnote{We could slow down the dynamics of ICT capital further, for example by assuming investment adjustment costs. Obviously, this would improve the fit of the model since it would allow the externality effect to play out even slower. Our key point here, however, is to show how the spillover is necessary to match the main features of the empirical responses. Without it, the model has no chance to capture the lagged effect of ICT shocks on TFP.}

More importantly, there is important economic intuition underlying our assumption of the spillover. Since the purpose of ICT goods is to enhance communication, the quality and speed of information acquisition and sharing is also determined by the overall diffusion of ICT goods. As a simple example, assume that in 1995 a scholar decides to purchase a computer with an Internet subscription in order to speed up the acquisition of information. She is very satisfied by this investment and does not regret the amount paid. However, at the same time, a former colleague who is now working in a different university decides to buy a computer with an Internet subscription as well. Once she learns of this development, the initial researcher decides to henceforth use her computer to exchange research material with her former colleague as well. They eventually decide to start a new project together, since being in two different universities is no longer a constraint for their joint research. None of the two scholars paid for this additional benefit when they purchased a computer, since they bought it for their personal use and did not expect the other to simultaneously acquire one also. What is actually happening here is that the research production of both researchers has been boosted by the positive externality that also the other colleague is endowed with the same technology. This externality is what we understand under the general-purpose nature of ICT goods and what our model captures with the assumption of the spillover.

The impulse response matching exercise of Section \ref{section:IRmatching} shows that a two-sector model featuring ICT capital and hard capital can only successfully match the empirical responses to an ICT shock if indeed the spillover effect is in action. Thus the main takeaway of this paper is that the interpretation of ICT as a general-purpose technology of the US between 1989-2017 is a good explanation of medium-run TFP growth dynamics. This relies on understanding the general-purpose nature of ICT goods as the extent to which their diffusion in the economy matters.

%As a result, we expect that ICT capital should display two main features. First of all, the effectiveness of an ICT good must be related to its own diffusion as captured by the spillover effect. Moreover, the cost of ICT items should not initially reflect an unexpected increase in their diffusion as pointed out by the assumption that the first feature is not internalized.

@@

In this paper, we look at the effect of ICT shocks on medium-run TFP from both an empirical and a theoretical perspective. In the first part of the paper, we investigate the effect of technological shocks specific to the ICT sector on future TFP. Using Structural VAR techniques on quarterly aggregate U.S. data from 1989, we impose two identifying restrictions to correctly identify ICT shocks: (i) they must be orthogonal to current productivity and (ii) maximize the impact response of ICT investment. In response to an ICT shock, ICT investment increases on impact and stays significant for more than 10 years; relative prices persistently decline as an additional support for the reliability of our identification strategy; and TFP rises after three quarters and remains significant for more than 10-years. In addition, variance decomposition analysis suggests that the quantitative impact of ICT shocks on TFP is economically significant since they explain about one third of the overall TFP fluctuations over a 10-year horizon.

In order to check whether our identification assumptions may confound our shock with a news shock which contemporaneously boosts ICT investment, we provide a series of different identification strategies and show that they deliver analogous results. Our main robustness check is a sequential identification of news shocks first and then ICT shocks. Encouragingly, neither this, nor any other robustness check affects our initial estimates. We view this as a confirmation of the causality relation which goes from current ICT investment to future TFP.

In the second part of the paper, we dig deeper into the nature of the link between current ICT investment and future TFP. We rationalize our empirical results by analyzing a two-sector DSGE model \`a la \cite{greenwood1997long} augmented by two main assumptions: (i) the sector with the faster technological improvement produces specifically ICT capital, and (ii) both production functions are augmented with a spillover related to the overall level of ICT capital which is not internalized by the agents. In particular, this last assumption points out our interpretation of the general-purpose nature of ICT capital: the effectiveness of ICT capital is enhanced by its diffusion, and the cost of ICT goods does not reflect this indirect benefit. We use both empirical and theoretical results to estimate main parameters of the model via impulse-response matching to an ICT technology shock. Results show that the parameter which governs the size of the spillover effect is positive and large, suggesting that our interpretation of the general-purpose nature of ICT capital is confirmed by the data. Thus, our investigation suggests not only that ICT is an important driver of medium-run TFP, but specifically that what matters for the evolution of TFP is whether ICT can act as a general-purpose technology. In particular, large externalities from using ICT goods can explain why a sector comprising a tiny fraction of both output and investment can play an important role for future productivity developments.
